Tempus: Probabilistic Network Latency Verifier

### 2022-10-20 paper/main.tex
[Paper] completed latency modelling chapter

\subsection{Latency Distribution}
Up to this point, we have talked about measuring the latency of a single packet by figuring out 
its path; analyzing which exact components it has traversed through.
However, when we try to generalize this framework and ask questions about the latency of multiple 
packets, it is apparent that the path alone is not a determinative information, as the latency 
of two packets propagating through the same path might be different due to a multitude of factors.

@@

\subsection{Path Decomposition and Convolution}
Expand on these points:
\begin{itemize}
    \item How do we get the path latency distribution? Component-wise measurement data
    \item What measurement? link delay and queuing delay $\rightarrow$ not all delays are 
    significant
    \item What to do with it? Simulation $\rightarrow$ lack of error guarantee
    \item Convolution
    \item Independence?
\end{itemize}

\section{Encoding}
We will start with the assertion that latency is a property that only make sense 
after connectivity between two nodes has been established. 
In other words, if two nodes in a network aren't connected (due to physical failure 
or ACL policy), then the latency between them will \textit{always} be infinite.
Because of this, we divide the problem of latency verification into two parts: 
verifying that two nodes are reachable (\textit{functional} property) and only if 
the functional property is fulfilled, we would verify whether the latency between 
two nodes fulfilled some additional condition (\textit{temporal} property)

@@

On top of this topology, we could define additional information that would be used
by a routing protocol to determine the valid path(s) between two nodes 
$src, dst \in V_t$ given a particular link failure scenario. 
In OSPF for example, we define a function $w_{ospf}: E_t \rightarrow \mathbb{N}$ as the edge-label 
that represents the positive weight of a link.

\subsection{Latency Graph}
There are many source of latency in a network. 
The principal challenge for us, then, is to determine the appropriate level of abstraction; 
detailed enough in order to accomplish our verification goal correctly but 
abstract enough for the problem to remain tractable. 
We settle on modeling two significant sources of latency: propagation and 
queuing delay.

To take both of these forms of delay into account, we've decided to model the latency of each
components with a univariate continuous random variable.
For propagation delay, the semantic of this random variable is relatively straightforward: it is 
the distribution of latency that a given link will introduce.
For queuing delay however, this random variable represents the delay that a given queuing process 
will introduce, marginalized over various traffic pattern that a given network state might have 
resulted.

% TODO: change this to v_i
To convey this additional notion of latency distribution, we form a second edge-labeled directed 
graph (called a latency graph) $G_l = (V_l, E_l)$.
For each network node $v_i \in V_t$, we want to ``expand'' this node to be able to represent the 
inner working of the queuing process within each node.

\begin{itemize}
    \item Each $v_i$ is expanded into $v_i^{in}, v_i^{out-1}, ..., v_i^{out-n} \in V_l$ 
    where $n$ is the amount $v_i$'s outward neighbors.

    \item Every inward edge that $v_i$ has in $G_t$ will all get directed to $v_i^{in}$.

    \item Every outward edge that $v_i$ has in $G_t$ will be sourced from one of the 
    $v_i^{out-j}$ instead.

    \item Add additional edges $e_i^j$ that connects from $v_i^{in}$ to each $v_i^{out-j}$.

    \item Finally, we label each $e \in E_l$ with the random variable as the latency distribution.
\end{itemize}

Semantically, $v_i^{out-j}$ represents a port, normal links that are copied from $G_t$ represents 
a link that has a propagation delay, and newly introduced links $e_i^j$ represents an output queue 
in node $v_i$ that has a queuing delay.

@@

\subsubsection{Functional Verification}
Let a network state $s_i$ be defined as a 2-tuple $(U, D)$ where $U$ is a set of links that are 
alive / up and $D$ is a set of links that are dead / down.
We then define a set $S$ that represents all possible network states in a given topology. 
Our goal is to identify a subset of $S$ where each element is a network state that can connect 
the source and destination node given a convergent behavior of a routing protocol.

@@

\subsubsection{Temporal Verification}
Since we assign each links and output queue in the network with a certain latency distribution, 
computing the probability of the temporal property involves convolving the the paths we got from 
functional verification and measuring the statistical property of the resulting distribution.
However, while NetDice has shrunk the exploration space compared to brute-force iteration, for
the purpose of temporal verification, further optimization can be done.

>>>>

\subsection{Path Latency Distribution}
Up to this point, we have talked about measuring the latency of a single packet by figuring out 
its path; analyzing which exact components it has traversed through.
However, when we try to generalize this framework and ask questions about the latency of multiple 
packets, it is apparent that the path alone is not a determinative information, as the latency 
of two packets propagating through the same path might be different due to a multitude of factors.

@@

\subsection{Path Decomposition and Convolution}
The final question that we had to decide on was how do we actually model path latency distribution
based on real component measurements.
Considering the complexity of factors that might determine the path latency distribution and the 
availability of measurement data, we settle on the assumption that the path latency distribution 
is composed of multiple independent distribution that corresponds to the latency each components 
in the path might introduce.
%TODO: confirm about independence?

Since not all components in the path will introduce latency with a non-negligible value, we 
specifically choose to model two source of latency in the path's components which we deem 
significant: \textbf{link propagation latency} and \textbf{queuing latency}.
%TODO: expand on how to get the data later

For propagation delay, the semantic of this random variable is relatively straightforward: it is 
the distribution of latency that a given link will introduce.
For queuing delay however, this random variable represents the delay that a given queuing process 
will introduce, marginalized over various traffic pattern that a given network state might have 
resulted.

In order to obtain the overall path latency distribution from these per-component latency 
distributions, we do a convolution operations over all the relevant component's latency 
distributions.
Since not all distributions can be convolved in a closed-form fashion, we use a numerical 
convolution technique with a guaranteed error bound. 
We initially consider a Monte-Carlo simulation in order to approach numerical convolution, but 
the lack of a general technique to measure error made us opt for the formerly mentioned method.

\section{Encoding}
We divide the problem of latency verification into two parts: 
verifying that two nodes are reachable (\textit{functional} property) and only if 
the functional property is fulfilled, we would verify whether the latency between 
two nodes fulfilled some additional condition (\textit{temporal} property).

@@

\subsection{Routing Label}
On top of this topology, we could define additional information that would be used
by a routing protocol to determine the valid path(s) between two nodes 
$src, dst \in V_t$ given a particular link failure scenario. 
In OSPF for example, we define a function $w_{ospf}: E_t \rightarrow \mathbb{N}$ as the edge-label 
that represents the positive weight of a link.

\subsection{Latency Label}
Expand on these points
\begin{itemize}
    \item distributions as label 
    \item convolve the relevant output queue latency + link latency
\end{itemize}
% There are many source of latency in a network. 
% The principal challenge for us, then, is to determine the appropriate level of abstraction; 
% detailed enough in order to accomplish our verification goal correctly but 
% abstract enough for the problem to remain tractable. 
% We settle on modeling two significant sources of latency: propagation and 
% queuing delay.

% To take both of these forms of delay into account, we've decided to model the latency of each
% components with a univariate continuous random variable.

% % TODO: change this to v_i
% To convey this additional notion of latency distribution, we form a second edge-labeled directed 
% graph (called a latency graph) $G_l = (V_l, E_l)$.
% For each network node $v_i \in V_t$, we want to ``expand'' this node to be able to represent the 
% inner working of the queuing process within each node.

% \begin{itemize}
%     \item Each $v_i$ is expanded into $v_i^{in}, v_i^{out-1}, ..., v_i^{out-n} \in V_l$ 
%     where $n$ is the amount $v_i$'s outward neighbors.

%     \item Every inward edge that $v_i$ has in $G_t$ will all get directed to $v_i^{in}$.

%     \item Every outward edge that $v_i$ has in $G_t$ will be sourced from one of the 
%     $v_i^{out-j}$ instead.

%     \item Add additional edges $e_i^j$ that connects from $v_i^{in}$ to each $v_i^{out-j}$.

%     \item Finally, we label each $e \in E_l$ with the random variable as the latency distribution.
% \end{itemize}

% Semantically, $v_i^{out-j}$ represents a port, normal links that are copied from $G_t$ represents 
% a link that has a propagation delay, and newly introduced links $e_i^j$ represents an output queue 
% in node $v_i$ that has a queuing delay.

@@

\subsection{Functional Verification}
Let a network state $s_i$ be defined as a 2-tuple $(U, D)$ where $U$ is a set of links that are 
alive / up and $D$ is a set of links that are dead / down.
We then define a set $S$ that represents all possible network states in a given topology. 
Our goal is to identify a subset of $S$ where each element is a network state that can connect 
the source and destination node given a convergent behavior of a routing protocol.

@@

\subsection{Temporal Verification}
Since we assign each links and output queue in the network with a certain latency distribution, 
computing the probability of the temporal property involves convolving the the paths we got from 
functional verification and measuring the statistical property of the resulting distribution.
However, while NetDice has shrunk the exploration space compared to brute-force iteration, for
the purpose of temporal verification, further optimization can be done.
